---
title: 'Homework 4: Tidy data and joins'
author: "Penny Kahn"
output: 
  html_document:
    keep_md: TRUE
    theme: yeti
    toc: TRUE
    toc_depth: 2
    toc_float: TRUE
---

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(DT))
```

#Exercise 1: Univariate Data Reshaping
You are expected to use pivot_wider() and pivot_longer() for reshaping, and ggplot2 for the plot.

## **Univariate Option 1**
_First let's remind ourselves of the structure of the gapminder dataset_
```{r}
head(gapminder)
```

### 1. Make a tibble with one row per year, and columns for life expectancy for two or more countries.

_Okay, so first I'm going to select columns country, year, and life expectancy. Then I'll filter the rows to one country of my choice from each continent. This will be the tibble I pipe into pivot_wider()_

```{r}
long <- gapminder %>%
  select(country, year, lifeExp) %>% 
  filter(country %in% c("Germany", "Rwanda", "Japan", "Bolivia", "New Zealand"))

long %>% datatable()
```

_Now we can widen my "long" tibble_
```{r}
wide <- long %>% 
  pivot_wider(id_cols = year,
              names_from = country,
              values_from = lifeExp)

wide %>% datatable()
```

### 2. Take advantage of this new data shape to scatterplot life expectancy for one country against that of another.

_Graphing the life expectancies of two countries against each other in this way allows us to compare their relative patterns over time. For example, We can see that Japan and Germany have both had steady increases in life expectancy over time which creates a positive linear relationship. Whereas, Germany and Rwanda have had quite different trajectories in life expectancy as there seems to be no linear relationship._
```{r}
wide %>% 
  ggplot(aes(x=Japan, y=Germany))+
    geom_point()+
    theme_bw()+
    labs(x="Japan life expectancy", y="Germany life expectancy")
```
```{r}
wide %>% 
  ggplot(aes(x=Rwanda, y=Germany))+
    geom_point()+
    theme_bw()+
    labs(x="Rwanda life expectancy", y="Germany life expectancy")
```

### 3. Re-lengthen the data
```{r}
wide_to_long <- wide %>% 
  pivot_longer(cols = -year,
               names_to = "country",
               values_to = "lifeExp")

wide_to_long %>% datatable()
```

## **Univariate Option 2**
_I have time, so I'm going to do this option as well for the practice._

### 1. Compute some measure of life expectancy (mean? median? min? max?) for all possible combinations of continent and year. Reshape that to have one row per year and one variable for each continent. Or the other way around: one row per continent and one variable per year.

_Again we'll look at the structure of gapminder_
```{r}
head(gapminder)
```

_And now I'm making a new column for the mean life expectancy in each continent in each year._
```{r}
long2 <- gapminder %>%
  group_by(continent, year) %>% 
  summarise(avg_life = mean(lifeExp))

long2 %>% datatable()
```

_And now to widen the data by having a column for each continent's average life expectancy for every year._
```{r}
wide2 <- long2 %>% 
  pivot_wider(id_cols = year,
              names_from = continent,
              values_from = avg_life)

wide2 %>% datatable()
```

### 2. Is there a plot that is easier to make with the data in this shape versus the usual form? Try making such a plot!

_With the data in this format, we can do a very simple ggplot to view the life expectancy patterns without having to filter or wrangle the data at all._
```{r}
wide2 %>% 
  ggplot(aes(x=year, y=Africa))+
    geom_point()+
    geom_line()+
    theme_bw()+
    labs(x="Year", y="Average life expectancy in Africa")
```

### 3. Re-lengthen the data.
```{r}
wide2_to_long2 <- wide2 %>% 
  pivot_longer(cols = -year,
               names_to = "continent",
               values_to = "avg_life")

wide2_to_long2 %>% datatable()
```

#Exercise 2: Multivariate Data Reshaping

Don’t worry about producing a plot here. You are expected to use pivot_wider() and pivot_longer() for reshaping.

## **Multivariate Option 1**
### 1. Make a tibble with one row per year, and columns for life expectancy and GDP per capita (or two other numeric variables) for two or more countries.

_I would like to put the country before the other variable in the name, but I can't figure out how to do that._
```{r}
wide3 <- gapminder %>% 
  select(country, year, lifeExp, gdpPercap) %>% 
  filter(country %in% c("Vietnam", "Chad", "Serbia")) %>% 
  pivot_wider(id_cols = year,
              names_from = country,
              names_sep = "_",
              values_from = c(lifeExp, gdpPercap))

wide3 %>% datatable()
```

### 2. Re-lengthen the data.
```{r}
wide3_to_long3 <- wide3 %>%
  pivot_longer(cols = -year,
               names_to = c(".value", "country"),
               names_sep = "_")

wide3_to_long3 %>% datatable()
```

## **Multivariate Option 2**
_Again, I have time so I'm going to do this for practice._

### 1. Compute some measure of life expectancy and GDP per capita (or two other numeric variables) (mean? median? min? max?) for all possible combinations of continent and year. Reshape that to have one row per year and one variable for each continent-measurement combination. Or the other way around: one row per continent and one variable for each year-measurement combination.

_I'm adding columns for the average life expecatncy and average GDP per capita for every continent in every year._
```{r}
long4 <- gapminder %>% 
  group_by(continent, year) %>% 
  summarise(avgLife = mean(lifeExp), avgGDP = mean(gdpPercap))

long4 %>% datatable()
```

_And now I'm widening it so that there are different columns for the averages in every year for each continent._
```{r}
wide4 <- long4 %>% 
  pivot_wider(id_cols = continent,
              names_from = year,
              names_sep = "_",
              values_from = c(avgLife, avgGDP))

wide4 %>% datatable()
```

### 2. Re-lengthen the data.
```{r}
wide4_to_long4 <- wide4 %>%
  pivot_longer(cols = -continent,
               names_to = c(".value", "year"),
               names_sep = "_")

wide4_to_long4 %>% datatable()
```

#Exercise 3: Table Joins

```{r}
guest <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/attend.csv")
email <- read_csv("https://raw.githubusercontent.com/STAT545-UBC/Classroom/master/data/wedding/emails.csv")
```

_Let's look at the guest and email datasets first._
```{r}
head(guest)
head(email)
```


### 3.1 For each guest in the guestlist (guest tibble), add a column for email address, which can be found in the email tibble.

_I am renaming the "guest" column to "name" in the email dataset so it matches. I'm also separating the names from each party into different rows. I asked Vincenzo in class when you would ever even use a right_join, so I guess this is a perfect example! The answer is: when you want to do manipulations to the table that isn't your base table before joining them._
```{r}
email %>% 
  rename(name = guest) %>% 
  separate_rows(name, sep = ", ") %>% 
  right_join(guest, by = "name") %>% 
  datatable()
```

### 3.2 Who do we have emails for, yet are not on the guestlist?
```{r}
email %>% 
  rename(name = guest) %>% 
  separate_rows(name, sep = ", ") %>%
  anti_join(guest, by = "name") %>% 
  datatable()
```

###3.3 Make a guestlist that includes everyone we have emails for (in addition to those on the original guestlist).
```{r}
email %>% 
  rename(name = guest) %>% 
  separate_rows(name, sep = ", ") %>%
  full_join(guest, by = "name") %>% 
  arrange(email) %>% 
  datatable()
```