---
title: 'Homework 1: Exploring Gapminder'
author: "Penny Kahn"
date: "9/12/2019"
output:
  beamer_presentation: default
  ioslides_presentation:
    fig_height: 4
    fig_width: 6
    smaller: yes
    widescreen: yes
---

```{r load packages, echo=FALSE}
library(tibble)
library(gapminder)
library(DT)
library(ggplot2)
```

## Intro to gapminder {.smaller}
This is the gapminder dataset. It contains information on the demographics of different countries through time.
```{r view}
datatable(as_tibble(gapminder))
```

## Exploring variables
<font size="3">We can create lists of the observations within different variables.</font>
```{r}
levels(gapminder$continent)
```
<font size="3">And find out counts of unique observations within variables. This function shows the number of entries in the dataset from each continent.</font>
```{r}
table(gapminder$continent)
```

## Plotting data {.smaller}
We can also make plots to explore the dataset
```{r}
ggplot(gapminder, aes(x=reorder(continent, lifeExp, FUN = median), y=lifeExp))+
  geom_boxplot()+
  labs(y="Life Expectancy", x="Continent")
```

## Plotting more data {.smaller}
```{r}
ggplot(subset(gapminder, continent=="Asia"), aes(x=reorder(country, lifeExp, FUN = median), y=lifeExp))+
  geom_boxplot()+
  coord_flip()+
  labs(y="Life Expectancy", x="Country")
```

