---
title: "Hw02_dplyr"
author: "Penny Kahn"
date: "9/20/2019"
output: html_document
---

```{r}
library(dplyr)
library(gapminder)
```


# Homework 2: Exploring gapminder using dplyr

## Exercise 1: Basic dplyr

1.1 Use filter() to subset the gapminder data to three countries of your choice in the 1970’s.
```{r}
gapminder %>%
  filter(country == "Canada" | country == "Bolivia" | country == "Turkey") %>% 
  filter(year == 1972 | year == 1977)
```

1.2 Use the pipe operator %>% to select “country” and “gdpPercap” from your filtered dataset in 1.1.
```{r}
gapminder %>%
  filter(country == "Canada" | country == "Bolivia" | country == "Turkey") %>% 
  filter(year == 1972 | year == 1977) %>% 
  select(country, gdpPercap)
```

1.3 Filter gapminder to all entries that have experienced a drop in life expectancy. Be sure to include a new variable that’s the increase in life expectancy in your tibble. Hint: you might find the lag() or diff() functions useful.
```{r}

```

1.4 Choose one of the following:

Filter gapminder so that it shows the max GDP per capita experienced by each country. Hint: you might find the max() function useful here.
OR

Filter gapminder to contain six rows: the rows with the three largest GDP per capita, and the rows with the three smallest GDP per capita. Be sure to not create any intermediate objects when doing this (with, for example, the assignment operator). Hint: you might find the sort() function useful, or perhaps even the dplyr::slice() function.
```{r}

```

1.5 Produce a scatterplot of Canada’s life expectancy vs. GDP per capita using ggplot2, without defining a new variable. That is, after filtering the gapminder data set, pipe it directly into the ggplot() function. Ensure GDP per capita is on a log scale.
```{r}

```

## Exercise 2: Explore individual variables with dplyr

Pick one categorical variable and one quantitative variable to explore. Answer the following questions in whichever way you think is appropriate, using dplyr:

What are possible values (or range, whichever is appropriate) of each variable?
What values are typical? What’s the spread? What’s the distribution? Etc., tailored to the variable at hand.
Feel free to use summary stats, tables, figures.
```{r}

```

## Exercise 3: Explore various plot types

Make two plots that have some value to them. That is, plots that someone might actually consider making for an analysis. Just don’t make the same plots we made in class – feel free to use a data set from the datasets R package if you wish.

A scatterplot of two quantitative variables.
One other plot besides a scatterplot.
You don’t have to use all the data in every plot! It’s fine to filter down to one country or a small handful of countries.
```{r}

```

## Recycling (Optional)
For people who want to take things further.

Evaluate this code and describe the result. Presumably the analyst’s intent was to get the data for Rwanda and Afghanistan. Did they succeed? Why or why not? If not, what is the correct way to do this?

filter(gapminder, country == c("Rwanda", "Afghanistan"))
Read What I do when I get a new data set as told through tweets from SimplyStatistics to get some ideas!
```{r}

```

## Tibble display (Optional)
Present numerical tables in a more attractive form using knitr::kable() for small tibbles (say, up to 10 rows), and DT::datatable() for larger tibbles.
```{r}

```

