---
title: 'Homework 3: dplyr and ggplot continued'
author: "Penny Kahn"
output: 
  html_document:
    keep_md: TRUE
    theme: yeti
---
```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(ggthemes))
suppressPackageStartupMessages(library(DT))
```

# Instructions

Pick three of the six tasks below, and produce:

* a tibble, using dplyr as your data manipulation tool;
* an accompanying plot of data from the tibble, using ggplot2 as your visualization tool; and
* some dialogue about what your tables/figures show (doesn’t have to be much).

## _Task Option 2_
### Get the maximum and minimum of GDP per capita for all continents.
```{r}
gapminder %>%
  group_by(continent) %>% 
  summarize(minGDP=min(gdpPercap), maxGDP=max(gdpPercap)) %>% 
  datatable
```
Wait, let's reshape that data to make it easier to graph:
```{r}
library(data.table)
```
```{r}
wide<-gapminder %>%
  group_by(continent) %>% 
  summarize(minGDP=min(gdpPercap), maxGDP=max(gdpPercap))

long <- melt(wide, id.vars="continent", variable.names="year")
```

I chose to do a very simple stripchart to show the differences in ranges of GDP per capita between the continents
```{r}
long %>% 
  ggplot(aes(x=continent, y=value))+
    geom_point()+
    geom_line()+
    labs(x="Continent", y="Range of GDP per capita")+
    theme_bw()
```



## _Task Option 5_
### How is life expectancy changing over time on different continents?

Here I have created a new column with average life expenctancy for each continent in each year.
```{r}
gapminder %>% 
  group_by(continent, year) %>% 
  mutate(avglife = mean(lifeExp)) %>% 
  distinct(continent, year, avglife) %>% 
  datatable()
```
I guess instead of making a new column and filtering out unique observations, we can use the summarize function:
```{r}
gapminder %>% 
  group_by(continent, year) %>% 
  summarize(avglife = mean(lifeExp)) %>% 
  datatable()
```

```{r}
gapminder %>% 
  group_by(continent, year) %>% 
  summarize(avglife = mean(lifeExp)) %>% 
  ggplot(aes(x=year, y=avglife, color=continent))+
    geom_line()+
    labs(x="Year", y="Average life expectancy")+
    theme_bw()
```

I tried graphing this without the summarize function but couldn't get it to work:
```{r}
gapminder %>% 
  ggplot(aes(x=year, y=lifeExp, color=continent))+
    geom_line(aes(group=continent))+
    labs(x="Year", y="Average life expectancy")+
    theme_bw()
```

I feel like this should work...


## _Task Option 6_
### Find countries with interesting stories. Open-ended and, therefore, hard. Promising but unsuccessful attempts are encouraged. This will generate interesting questions to follow up on in class.

Since we've been working with gapminder, I've noticed a huge drop in life expectancy in african countries in the early nineties. I'd like to look at that more clearly. First let's look at the countries in Africa:
```{r}
gapminder %>% 
  filter(continent == "Africa") %>% 
  ggplot(aes(x=year, y=lifeExp))+
    geom_line(aes(group=country))+
    theme_bw()
```

Let's zoom in on the 90s
```{r}
gapminder %>% 
  filter(continent == "Africa") %>% 
  filter(year>1985) %>% 
  ggplot(aes(x=year, y=lifeExp))+
    geom_line(aes(group=country))+
    theme_bw()
```

Looks like there's a few that have more "significant" drops than others. Let's tease out just the countries that have a "big" drop at some point
```{r}
gapminder %>% 
  filter(continent == "Africa") %>% 
  mutate(lifeExp_inc = lifeExp - lag(lifeExp)) %>% 
  filter(lifeExp_inc < -5, year !=1952) %>% 
  datatable()
```

Now let's plot just those countries. I'm sure there's a more elegant way to do this, but for now I'll stick to what I know :)
```{r}
gapminder %>% 
  filter(country=="Botswana"|country=="Lesotho"|country=="Liberia"|country=="Namibia"|country=="Rwanda"|country=="South Africa"|country=="Swaziland"|country=="Zambia"|country=="Zimbabwe") %>% 
  ggplot(aes(x=year, y=lifeExp, color=country))+
    geom_line(aes(group=country))+
    theme_bw()
```

So what happened to these countries from the late 80s to now? All except Liberia are located in the southern half of Africa. I'm guessing it was the HIV/AIDS epidemic. 

I would love to make a graph with the lines for all the countries in the world in black except for these 9 countries in the colors shown above. I couldn't figure it out though

Here's a table to go along with it. It shows the change in life expectancy from year to year in order to highlight the big drops:
```{r}
gapminder %>% 
  filter(country=="Botswana"|country=="Lesotho"|country=="Liberia"|country=="Namibia"|country=="Rwanda"|country=="South Africa"|country=="Swaziland"|country=="Zambia"|country=="Zimbabwe") %>% 
  mutate(lifeExp_inc = lifeExp - lag(lifeExp)) %>% 
  select(country,year,lifeExp, lifeExp_inc) %>% 
  datatable()
```

